Think deep for reef resilience

### 2025-07-10 paper/paper_draft.qmd
Add transparency and adjust plotting for carbonate budget plots

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Current modelling projects severe declines in coral cover under projections of Degree Heating Weeks [**Bozec 2025**]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

@@

While these environmental and biological factors determine whether individual corals survive thermal stress events, the long-term viability of coral reefs depends not only on coral survival but also on the maintenance of reef structure through positive carbonate budgets. Scleractinian corals are primary habitat structures on coral reefs, and the accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes including coral accumulation [**Perry**], bioerosion [**Tribollet**], physical damage generating coral rubble [**Kenyon**], and calcification and CaCO₃ dissolution rates [**Eyre 2014, 2018**]. The balance of these processes and occurrence of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity [**Perry, Desbiens**], rates of physical environmental processes, and ocean acidity [**Perry, Eyre**]. Total live coral cover on a reef highly influences the maintenance of net positive carbonate budgets, with estimates ranging from 10 to 20% [**Perry**] of a reef's total area required to maintain positive budgets. For mid and outer shelf reefs on the Great Barrier Reef, this threshold is approximately 17% live coral cover [**Desbiens**].

@@

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. It follows the reef definitions as provided by the Great Barrier Reef Marine Park Authority [https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::great-barrier-reef-features-/explore?location=-19.221541%2C145.776418%2C4.94]. Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6; **citation**) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon]. Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology [@liu2017; @skirving2020]. These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration [@forster2021]. The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020; IPCC -@climate2023].

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals growing in diameter and competing for limited space over time [@ribeirodealmeida2024]. This model incorporates linear extension (growth) and background mortality rates that are dependent on the size and functional group of a coral colony. Environmental disturbances and reproduction are not directly considered within CoralBlox. Instead, these factors are considered by making changes to the distribution of colonies within CoralBlox timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform [@iwanaga2023]. Within ADRIA, we incorporate coral life history and environmental events in the following order: coral colony growth; coral reproduction; larval connectivity and settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled together to represent the system, we refer to both as ADRIA-CoralBlox.

@@

** need information on GBR1 connectivity used **

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we clustered reefs by their temporal behaviour as indicated by their coral cover over time. The Complexity-Invariant Distance (CID) between each timeseries was calculated to identify timeseries that behave similarly [@batista2014]. The CID corrects the Euclidean distance between pairs of timeseries (which accounts for magnitude differences) with the complexity of each timeseries (obtained through a measure of their variability) to ensure reefs are clustered based on their underlying temporal dynamics rather than simply their average coral cover. The process creates a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020].

Timeseries clustering analyses were performed at the reefal bioregion scale (defined in [https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::reefal-marine-bioregions/about]).
Additional analyses were also performed at larger spatial scales, such as the GBRMPA management regions and whole-of-GBR scale. These were found to lead to identical conclusions and so are not shown here.
Results of these analyses can be found at [**insert citation for repository**] for those interested.

Reefs in each timeseries cluster are compared to identify characteristics that differ between low, medium and high coral cover clusters. These characteristics include reef median depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation between total coral cover and inversed DHW timeseries, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and total strength of larval connections between reefs. To analyse the DHW - tolerance levels of reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep * taxa * reef * scenario matrix. The mean timeseries was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

@@

## Reef carbonate budget
The number of years reefs maintain coral cover above a required carbonate budget level substantially decreases as the required carbonate budget threshold (as a percentage of reef area) increases. Reefs with greater median depth levels maintain coral cover above required carbonate budget levels for longer than reefs at shallow depths [@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation with the number of years a reef maintains coral cover above the required threshold [@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation decreases as the required carbonate budget threshold increases. This is likely due to an increasing number of reefs being unable to maintain coral cover above threshold levels as the threshold increases. There are no clear relationships between reef total connectivity levels and the number of years reefs maintain coral cover above carbonate budget threshold levels [@sfig-access-esm1-5-conn-carbonate]. Reef log (natural) total connectivity strength has no correlation with the number of years a reef maintains cover above the required thresholds. Log connectivity strength - carbonate budget maintainence correlation slightly decreases as the required carbonate budget threshold increases.

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced, and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to calculate the degree of protection provided to a reef based on the median depth. Our results suggest that within shallow reef ecosystems (depth < 30m), depth contributes greatly to reef resilience and the ability of the reef to maintain cover under heat stress. While similar findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses have often focussed on the potential of mesophotic reefs (> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific.
<!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->

@@

## Model parameterisation

@@

![Mean timeseries DHW values for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean width=`{julia} violin_width`}
<!--
![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol width=`{julia} violin_width`} -->

>>>>

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Current modelling projects severe declines in coral cover under projections of Degree Heating Weeks \[**Bozec 2025**\]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

@@

While these environmental and biological factors determine whether individual corals survive thermal stress events, the long-term viability of coral reefs depends not only on coral survival but also on the maintenance of reef structure through positive carbonate budgets. Scleractinian corals are primary habitat structures on coral reefs, and the accumulation and erosion of coral reef carbonate frameworks is controlled by a range of processes including coral accumulation \[**Perry**\], bioerosion , physical damage generating coral rubble \[**Kenyon**\], and calcification and CaCO₃ dissolution rates \[**Eyre 2014, 2018**\]. The balance of these processes and occurrence of net carbonate framework production (referred to as a positive carbonate budget) are influenced by factors such as coral cover and complexity \[**Perry, Desbiens**\], rates of physical environmental processes, and ocean acidity \[**Perry, Eyre**\]. Total live coral cover on a reef highly influences the maintenance of net positive carbonate budgets, with estimates ranging from 10 to 20% \[**Perry**\] of a reef's total area required to maintain positive budgets. For mid and outer shelf reefs on the Great Barrier Reef, this threshold is approximately 17% live coral cover \[**Desbiens**\].

@@

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. 
It follows the reef definitions as provided by the Great Barrier Reef Marine Park Authority \[https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::great-barrier-reef-features-/explore?location=-19.221541%2C145.776418%2C4.94\]. Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6; **citation**) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon]. Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology [@liu2017; @skirving2020]. These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration [@forster2021]. The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020; IPCC -@climate2023].

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals growing in diameter and competing for limited space over time [@ribeirodealmeida2024]. Details on how coral demographic processes are implemented in ADRIA - CoralBlox can be found in the supplementary material \[**link ADRIA - CoralBlox**\].

@@

\*\* need information on GBR1 connectivity used \*\*

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we clustered reefs by their temporal behaviour as indicated by their coral cover over time. The Complexity-Invariant Distance (CID) between each timeseries was calculated to identify timeseries that behave similarly [@batista2014]. The CID corrects the Euclidean distance between pairs of timeseries (which accounts for magnitude differences) with the complexity of each timeseries (obtained through a measure of their variability) to ensure reefs are clustered based on their underlying temporal dynamics rather than simply their average coral cover. The process creates a distance matrix of n \* n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020].

Timeseries clustering analyses were performed at the reefal bioregion scale (defined in \[https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::reefal-marine-bioregions/about\]). Additional analyses were also performed at larger spatial scales, such as the GBRMPA management regions and whole-of-GBR scale. These were found to lead to identical conclusions and so are not shown here. Results of these analyses can be found at \[**insert citation for repository**\] for those interested.

Reefs in each timeseries cluster are compared to identify characteristics that differ between low, medium and high coral cover clusters. These characteristics include reef median depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation between total coral cover and inversed DHW timeseries, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and total strength of larval connections between reefs. To analyse the DHW - tolerance levels of reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep \* taxa \* reef \* scenario matrix. The mean timeseries was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

@@

## Reef carbonate budget

The number of years reefs maintain coral cover above a required carbonate budget level substantially decreases as the required carbonate budget threshold (as a percentage of reef area) increases. Reefs with greater median depth levels maintain coral cover above required carbonate budget levels for longer than reefs at shallow depths [@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation with the number of years a reef maintains coral cover above the required threshold [@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation decreases as the required carbonate budget threshold increases. This is likely due to an increasing number of reefs being unable to maintain coral cover above threshold levels as the threshold increases. There are no clear relationships between reef total connectivity levels and the number of years reefs maintain coral cover above carbonate budget threshold levels [@sfig-access-esm1-5-conn-carbonate]. Reef log (natural) total connectivity strength has no correlation with the number of years a reef maintains cover above the required thresholds. Log connectivity strength - carbonate budget maintainence correlation slightly decreases as the required carbonate budget threshold increases.

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced, and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to calculate the degree of protection provided to a reef based on the median depth. Our results suggest that within shallow reef ecosystems (depth \< 30m), depth contributes greatly to reef resilience and the ability of the reef to maintain cover under heat stress. While similar findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific. <!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->

@@

## ADRIA - CoralBlox

CoralBlox model incorporates linear extension (growth) and background mortality rates that are dependent on the size and functional group of a coral colony. Environmental disturbances and reproduction are not directly considered within CoralBlox. Instead, these factors are considered by making changes to the distribution of colonies within CoralBlox timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform [@iwanaga2023]. Within ADRIA, we incorporate coral life history and environmental events in the following order: coral colony growth; coral reproduction; larval connectivity and settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled together to represent the system, we refer to both as ADRIA-CoralBlox.

### Model parameterisation

@@

![Mean timeseries DHW values for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean width=`{julia} violin_width`} <!-- ![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol width=`{julia} violin_width`} --\>
